---
title: "dplyr Superhero"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
---

## Learning Goals  
*At the end of this exercise, you will be able to:*    
1. Develop your dplyr superpowers so you can easily and confidently manipulate dataframes.  

## Instructions
This is a challenging exercise that will give you a chance to practice your dplyr skills and add a few new tricks with `mutate()`. This is isn't easy, so if you get stuck make sure to post to the slack channel for help- that's what Wonder Woman would do...  

## Load the tidyverse and janitor
```{r}
library("tidyverse")
library("janitor")
```

## Load here
```{r}
library("here")
```

## Load the superhero data
These are data taken from comic books and assembled by fans. They include a good mix of categorical and continuous data. Data taken from: https://www.kaggle.com/claudiodavi/superhero-set  

Check out the way I am loading these data. If I know there are NAs, I can take care of them at the beginning. But, we should do this very cautiously. At times it is better to keep the original columns and data intact. Notice that I am also cleaning the names of the variables! The package `janitor` has a lot of superpowers.
```{r}
superhero_info <- read_csv(here("data", "heroes_information.csv"), na = c("", "-99", "-")) %>% clean_names()
```

```{r}
superhero_powers <-read_csv(here("data", "super_hero_powers.csv"), na = c("", "-99", "-")) %>% clean_names()
```

## `tabyl`
The `janitor` package has many awesome functions that we will explore. Here is its version of `table` which not only produces counts but also percentages. Very handy! Let's use it to explore the proportion of good guys and bad guys in the `superhero_info` data.  
```{r}
tabyl(superhero_info, alignment)
```

1. Notice that we have some neutral superheros! Who are they?
```{r}

```

## `superhero_info`
2. Let's say we are only interested in the variables name, alignment, and "race". How would you isolate these variables from `superhero_info`?
```{r}

```

## Not Human
3. List all of the superheros that are not human.
```{r}

```

## Good and Evil
4. Let's make two different data frames, one focused on the "good guys" and another focused on the "bad guys".  

Good  
```{r}

```

Bad  
```{r}

```

5. Among the good guys, Who are the Asgardians?
```{r}

```

6. Among the bad guys, who are the male humans over 200 inches in height?
```{r}

```

7. OK, so are there more good guys or bad guys that are bald (personal interest)?
```{r}

```

```{r}

```

8. Let's explore who the really "big" superheros are. In the `superhero_info` data, which have a height over 200 or weight over 300?
```{r}

```

## Height to Weight Ratio
It's easy to be strong when you are heavy and tall, but who is heavy and short? Which superheros have the highest height to weight ratio? In order to do this, we need to make a new variable in our dataframe. To do this, we use `mutate()`.
```{r}
superhero_info %>%
  mutate(height_weight_ratio=height/weight) %>%
  select(name, height_weight_ratio) %>%
  arrange(desc(height_weight_ratio))
```

Notice that mutate() uses regular mathematical expressions. One thing to keep in mind is that the new column is always added at the very end of your existing dataframe.  

9. Weight is currently represented in pounds and height is in inches. Use mutate() to add two new variables that convert these to kilograms and centimeters.  
```{r}

```

## `superhero_powers`
Have a quick look at the `superhero_powers` data frame.  
```{r}

```

10. How many superheros have a combination of accelerated healing, durability, and super strength?
```{r}

```

## `kinesis`
11. We are only interested in the superheros that do some kind of "kinesis". How would we isolate them from the `superhero_powers` data?
```{r}

```

12. Pick your favorite superhero and let's see their powers!
```{r}

```

## That's it!
You are now a dplyr superhero!