---
title: "Transforming data 2: `filter()`"
date: "`r Sys.Date()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

## Learning Goals  
*At the end of this exercise, you will be able to:*    
1. Use the functions of dplyr (filter, select, arrange) to organize and sort data frames.  

## Questions
If you have question or need help with a specific problem, please [email me](mailto: jmledford@ucdavis.edu).  

## Review  
In our previous meeting, we used `select()` to extract columns of interest from a data frame. This helps us focus our attention on the variables relevant to our question. However, it doesn't allow us to extract observations from within the data frame. The `filter()` function allows us to extract data that meet specific criteria. When combined with `select()`, we have the power to transform, shape, and explore data with the potential to make new discoveries.  

## Load the `tidyverse` and `janitor`
```{r message=FALSE, warning=FALSE}
library("tidyverse")
library("janitor")
```

## Load `here` separately so you know your directory  
```{r}
library("here")
```

## Load the data
For these exercises we will use the data from the previous meeting.
```{r}

```

## `filter()`
Unlike `select()`, `filter()` allows us to extract data that meet specific criteria within a variable. Let's say that we are only interested only in observations for the year 2019. We can use `filter()` to extract these observations.
```{r}
glimpse(edu_dat)
```

First question- what years are represented in the data?  
```{r}

```

```{r}
filter(edu_dat, year=="2019")
```

Similarly, if we are only interested in students with a final exam score greater than or equal to 70 we can use `filter()` to extract these observations.  
```{r}
filter(edu_dat, final_exam >= 70)
```

+ `filter()` allows all of the expected operators; i.e. >, >=, <, <=, != (not equal), and == (equal).  

## Using `filter()` on multiple conditions
You can also use `filter()` to extract data based on multiple conditions. Below we extract only the students from 2019 with a final exam score greater or equal to than 70.
```{r}
filter(edu_dat, year == 2019 & final_exam >= 70)
```

Notice that the `|` operator generates a different result.
```{r}
filter(edu_dat, year == 2019 | final_exam >= 70)
```

Rules:  
+ `filter(condition1, condition2)` will return rows where both conditions are met.  
+ `filter(condition1, !condition2)` will return all rows where condition one is true but condition 2 is not.  
+ `filter(condition1 | condition2)` will return rows where condition 1 and/or condition 2 is met.  
+ `filter(xor(condition1, condition2)` will return all rows where only one of the conditions is met, and not when both conditions are met.  

## Practice in the chat
Filter out the students from 2019 with a prior gpa greater than or equal to 2 AND a final exam score greater than 70.
```{r}

```

## Using `filter()` for multiple observations within the same variable 
Filtering multiple values within the same variable requires the `%in%` [operator](https://www.tutorialspoint.com/r/r_operators.htm).    
```{r}
filter(edu_dat, gpa_prior %in% c(2.5, 3.0))
```

Alternatively, you can use `between` if you are looking for a range of specific values.  
```{r}
filter(edu_dat, between(gpa_prior, 2.5, 3.0))
```

You can also extract observations "near" a certain value but you need to specify a tolerance.  
```{r}
filter(edu_dat, near(gpa_prior, 3, tol = 0.2))
```

## Practice
1. Transform the data to only include final exam scores for freshmen.
```{r}

```

Feeling fancy? Use the output to make a barplot that shows counts by final exam score.
```{r}

```

2. Do we have any struggling freshmen? Transform the data to only include freshmen with a gpa less 2.5.
```{r}

```

3. Filter the data to include transfers from 2016 with a gpa between 3.5-4.0.
```{r}

```

4. Your department chair is interested in a comparison of enrollment between class 1 and class 2 for each year represented in the data. How would you show this as a summary and on a plot?
```{r}

```

```{r}

```

## That's it!  
`filter()` is a powerful tool for reshaping your data by extracting values of interest. When used in combination with `select()`, you innumerable ways to transform your data.