---
title: "Day2_2activity"
author: "Emily Weigel"
date: "7/28/2021"
output: html_document
---


## Review
Now that you have been introduced to `ggplot`, let's practice. Remember that plots are built in layers: **plot= data + geom_ + aesthetics**. We have to specify each of these in order for a plot to be produced. If you get stuck, it is often helpful to stop and make a quick sketch of what you want or expect to see on a piece of scratch paper.  

Let's review using the `penguins` data. First, get an idea of the structure: Are the data tidy? Are there NA's? Are the variables discrete, categorical, or continuous?
```{r}
library(palmerpenguins)
library(tidyverse)
penguins
```

Let's examine how many penguins are on each island.
```{r}
penguins %>% 
  count(island, species, sort = F)
```

What if we wanted a plot that showed the number of measured penguins per species and the number of measured penguins per island?
```{r}
penguins %>% 
  group_by(island) %>% 
  ggplot(aes(x=species))+
  geom_bar()
```

```{r}
penguins %>% 
  group_by(species) %>% 
  ggplot(aes(x=island))+
  geom_bar()
```

How about average bill length by sex?
```{r}
penguins %>% 
  filter(sex!="NA") %>% 
  group_by(sex) %>% 
  summarise(ave_bill_length=mean(bill_length_mm))
```

```{r}
penguins %>% 
  filter(sex!="NA") %>% 
  group_by(sex) %>% 
  summarise(ave_bill_length=mean(bill_length_mm)) %>% 
  ggplot(aes(x=sex, y=ave_bill_length)) +
  geom_col()
```

```{r}
penguins
```

##Resources
- [ggplot2 cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
- [`ggplot` themes](https://ggplot2.tidyverse.org/reference/ggtheme.html)
- [Rebecca Barter `ggplot` Tutorial](http://www.rebeccabarter.com/blog/2017-11-17-ggplot2_tutorial/)

## That's it! Good work on this!